---
title: "PA1_template.Rmd"
author: "Carlos Rodriguez-Contreras"
date: "August 16, 2015"
output: 
  html_document: 
    keep_md: yes
---
# Data Science Specialization
## Johns Hopkins University - Coursera
### Reproducible Research
### Assessment 1 
Carlos Rodriguez-Contreras  

This R Markdown document corresponds to the first peer assessment of the course Reproducible Research.  

## Loading and preprocessing the data  

```{r loadingData}
datasetURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(datasetURL, destfile = "zippedFile.zip", method = "curl")

# Unzip the dataset into work directory:
unzip("zippedFile.zip")
list.files()
Activity.Monitoring.Data <- read.csv("activity.csv", header = TRUE, na.strings = "NA")
str(Activity.Monitoring.Data)
Activity.Monitoring.Data$date = as.Date(as.character(Activity.Monitoring.Data$date), format = "%Y-%m-%d")

# Eliminating NA values from whole dataset:
library(Hmisc)
AMDwoNA <- na.delete(Activity.Monitoring.Data)
head(AMDwoNA)
```

## Mean total numbers of steps taken per day  

The dataframe *AMDwoNA* is splitted by date, then for each day the number of steps recorded in 5 min intervals are accumulated.

```{r, preprocessing1}
daily <- split(AMDwoNA, AMDwoNA$date)
dailySteps <- lapply(daily, function(x) {sum(x$steps)} )
dailySteps <- as.numeric(dailySteps)
```

### Generation of the Histogram without Missing Values:

```{r histogram1}
opar <- par()
par(bg = "cornsilk")
hist(dailySteps, xlab = "Number of Steps", main = "Total Steps per Day removing NA", labels = TRUE, ylim = c(0, 30), col = "lavender")
```

### Computing the Mean and the Median of the steps without missing values:

```{r meanMedian}
paste("The Mean of the total number of steps taken per day is:", round(mean(dailySteps), 2))
paste("The Median of the total number of steps taken per day is:", round(median(dailySteps), 2))
```

## Average daily activity pattern

The dataframe *AMDwoNA* is splitted by interval, then for each interval the mean of steps recorded are computed.

```{r dailiActivity}
AMDwoNA$interval <- as.factor(AMDwoNA$interval)
interval <- split(AMDwoNA, AMDwoNA$interval)
intervalSteps <- lapply(interval, function(x) {mean(x$steps)} )
intervalSteps <- as.numeric(intervalSteps)
interval <- as.numeric(levels(AMDwoNA$interval))
```

### Generation of the Time Series plot of intervals - average of steps:

```{r timeSeriesPlot}
par(bg = "cornsilk")
plot(interval, intervalSteps, type = "l", main = "Average Daily Activity Pattern", xlab = "Intervals", ylab = "Average of steps by interval", col ="darkblue")
```

### Computing the interval with the maximum number of steps:

```{r maxSteps}
newTable <- as.data.frame(cbind(interval, intervalSteps))
mostActiveInterval <- newTable[which(newTable$intervalSteps == max(newTable$intervalSteps)), ]
paste ("The most active interval is", mostActiveInterval$interval, "with", mostActiveInterval$intervalSteps)
```

## Imputing missing values

Identifying the NA (missing values) from the original data set *Activity.Monitoring.Data* computing the ammount of NA and impute them with the mean of the interval.

```{r missingValues}
AMDimputedNA <- Activity.Monitoring.Data
missingValues <- is.na(AMDimputedNA$steps)
paste("The total number of missing values is", sum(missingValues))
AMDimputedNA[missingValues, ]$steps <- intervalSteps
```

### Generating a Histogram with NA filled in:

```{r histogram2}
daily2 <- split(AMDimputedNA, AMDimputedNA$date)
dailySteps2 <- lapply(daily2, function(x) {sum(x$steps)} )
dailySteps2 <- as.numeric(dailySteps2)

# Creating the histogram for Total Steps per Day:
opar <- par()
par(bg = "cornsilk")
hist(dailySteps2, xlab = "Number of Steps", main = "Total Steps per Day imputing NA", labels = TRUE, ylim = c(0, 40), col = "lavender")
```

### Computing the Mean and the Median of the steps with missing values:

```{r meanMedian2}
paste("The Mean of the total number of steps taken per day is:", round(mean(dailySteps2), 2))
paste("The Median of the total number of steps taken per day is:", round(median(dailySteps2), 2))
```

It is noticed that there is a little impact of imputing missing data, this is in part for being 10% approx. This also means that using the mean as imputation method is quite acceptable.

## Differences in activity patterns between weekdays and weekends

For this to be computed, it is used the *weekdays()* function, in this particular case, names of days are displayed in Spanish and so, the computations used special characters (UTF-8 set of characters). Anyway, the results must be correct.

```{r weekdays}
daysOfWeek <- as.list(weekdays(AMDimputedNA$date))
TypeOfDays <- lapply(daysOfWeek, function(x){
    if (x != "sábado" & x != "domingo") x <- "Weekday"
    else x <- "Weekend"
})

AMDimputedNA$type <- as.factor(as.character(TypeOfDays))
```

### Genaration of plots using ggplot2:

```{r weekdaysGraph}
library(ggplot2)
qplot(interval, steps, data = AMDimputedNA, geom = "line", facets = type ~ . , main = "Comparison between Weekdays and Weekends")
```

It is noticed that during Weekends the steps pattern is more uniform which means more people are involved in some kind of fitness activities.
